### 2013-03-28 doc/sosp13/tasks.tex

The basic unit of execution in \Grappa is a {\em task}. Each hardware core has
a single operating system thread pinned to it. When tasks are ready to
execute, they are mapped to a {\em worker}, which is akin to a user-level
thread. \Grappa keeps a large number of workers per hardware core, this is the
core reason why \Grappa is able to tolerate latencies well. Below we discuss
the implementation of our task management support and then describe how
applications expose parallelism to the \Grappa runtime.

@@

\paragraph{Tasks}
Each task is represented by a function pointer and its arguments, represented
as 32-byte entities: a 64-bit function pointer plus three 64-bit arguments. We
use three arguments because tasks are often generated via a parallel loop
decomposition, in which each task needs three kinds of data. Thus, tasks
include:

\noindent {\it function pointer}, or the addresses of the routine to run ---
we configure the system nodes to disable address randomization, so that
function pointers are valid across process images;

@@

\noindent {\it synchronization argument}, often used to determine when all
tasks that are part of a loop have finished, in the form of q global pointer
to a synchronization object allocated at the core that spawned a group of
tasks.

@@

\paragraph{Workers}
When a new task is spawned, it is enqueued until it is ready for execution.
Tasks are not allocated any execution resources (a worker) until the scheduler
decides to run them. A worker is simply a collection of status bits and a
stack, allocated at a particular core. There are two types of workers: (1) user worker; and (2) system worker.

User workers execute application tasks. Once a task is mapped to a worker it
stays with that worker until it completes. System workers execute runtime
management activities such as communication and statistics collection. User
workers and system workers might have different scheduling priorities. System
workers execute periodically and in idle times, to guarantee timely
communication services.

\paragraph{Scheduling} 
During execution, a worker yields control of its core whenever it performs a
long-latency operation, allowing the processor to remain busy while waiting
for the operation to complete. In addition, a programmer can direct scheduling
explicitly via the \Grappa API calls shown in Figure~\ref{fig:scheduling}. To
minimize yield overhead, the \Grappa scheduler operates entirely in user-space
and does little more than store register state of one task and load that of
another.

@@

Each core in a \Grappa system has its own independent scheduler. Schedulers
have three main responsibilities: servicing system workers that process
communication requests as described in Section~\ref{sec:communication};
rescheduling workers that have been waiting on long-latency operations; and
assigning ready tasks to worker resources that have become idle.

Each scheduler has four queues:

\noindent {\it ready worker queue}, a FIFO queue of tasks that are
  matched with workers and are ready to execute;

@@

% Pretty sure this is not true anymore.  Even if it is, it's not important -Mark
%
%Since the compiler sees all calls to the context switch routine, we
%can save even less state. Our context switch routine appears to the
%compiler as inline assembly; we declare all the registers we need
%to save as ``clobbered'' by the inline assembly routine, and the
%compiler will issue its own save and restore code as needed. This allows the
%compiler to avoid saving any registers that are not used, or are used
%for temporary values that are not needed after the context switch.

@@

\paragraph{Work stealing} 
When the scheduler finds no work to assign to its workers, it commences to
steal tasks from other cores using an asynchronous \texttt{call\_on} active
message. It chooses a victim at random until it finds one with a non-zero
amount of work in its public task queue. The scheduler steals half of the
tasks it finds at the victim. Work stealing is particularly interesting in
\Grappa since performance depends on having many active worker threads on each
core. Even if there are many active threads, if they are all suspended on
long-latency operations, then the core is underutilized. We choose to initiate
one outstanding steal whenever the ready queue is empty there are idle worker
threads.

@@

\Grappa provides four methods for expressing parallelism, shown in
Figure~\ref{fig:expressing-parallelism}. First, when the programmer identifies
work that can be done in parallel, the work may be wrapped up in a function
and queued with its arguments for later execution using a \texttt{spawn}.
Second, a programmer can use \texttt{spawn\_on} to spawn a task on a specific
core in the system or at the home core of a particular memory location. Third,
the programmer can invoke a parallel for loop, provided that the trip count is
known at loop entry. The programmer specifies a function pointer along with
start and end indices and an optional threshold to control parallel overhead.
\Grappa does {\em recursive decomposition} of iterations, similar to Cilk's
cilk\_for construct~\cite {cilkforimplementation}, and TBB's {\tt
parallel\_for}~\cite{intel_tbb}. It generates a logarithmically-deep tree of
tasks, stopping to execute the loop body when the number of iterations is
below the required threshold. Fourth, a programmer may want to execute an active message; that is, to run a
small piece of code on a particular core in the system without waiting for
execution resources to be available. For example, custom synchronization
primitives execute this way, as a function executed on the core where the data
lives. \Grappa provides the \texttt{call\_on} call for this purpose.

>>>>

The basic unit of execution in \Grappa is a {\em task}. Each hardware core has a single operating system thread pinned to it. When tasks are ready to execute, they are mapped to a {\em worker}, which is akin to a user-level thread. \Grappa keeps a large pool of workers per hardware core. Below we discuss the implementation of our task management support and then describe how applications expose parallelism to the \Grappa runtime.

@@

\paragraph{Tasks} Each task is represented by a 64-bit function pointer and three 64-bit arguments.  We use three arguments because tasks are often generated via a parallel loop decomposition, in which each task needs three kinds of data:

@@

\noindent {\it synchronization argument}, often used to determine when all
tasks that are part of a loop have finished, in the form of a global pointer
to a synchronization object allocated at the core that spawned a group of
tasks.

@@

\paragraph{Workers} Workers execute application tasks.  A worker is simply a collection of status bits and a stack, allocated at a particular core.  When a task is ready to be executed it is assigned to a worker.  Once a task is mapped to a worker it stays with that worker until it completes.

%% Note sure this is even true.  And it's confusing and not relevant regardless. -M
% There are two types of workers: (1) user worker; and (2) system worker. 
%System workers execute runtime management activities such as communication and statistics collection. User workers and system workers might have different scheduling priorities. System workers execute periodically and in idle times, to guarantee timely communication services.

\paragraph{Scheduling} 
During execution, a worker yields control of its core whenever it performs a
long-latency operation, allowing the processor to remain busy while waiting
for the operation to complete. In addition, a programmer can direct scheduling
explicitly via the \Grappa API calls shown in Figure~\ref{fig:scheduling}. To
minimize yield overhead, the \Grappa scheduler operates entirely in user-space
and does little more than store register state of one worker and load that of
another.

@@

Each core in a \Grappa system has its own independent scheduler.  The scheduler has a single queue of active workers:

\noindent {\it ready worker queue}, a FIFO queue of workers that are
  matched with workers and are ready to execute;

Each scheduler also has three queues of tasks waiting to be assigned a worker:

@@

@@

\paragraph{Work stealing} 
When the scheduler finds no work to assign to its workers, it commences to
steal tasks from other cores using an asynchronous \texttt{call\_on} active
message. It chooses a victim at random until it finds one with a non-zero
amount of work in its public task queue. The scheduler steals half of the
tasks it finds at the victim. Work stealing is particularly interesting in
\Grappa since performance depends on having many active worker threads on each
core. Even if there are many active threads, if they are all suspended on
long-latency operations, then the core is underutilized.

@@

\Grappa provides four methods for expressing parallelism, shown in
Figure~\ref{fig:expressing-parallelism}. First, when the programmer identifies
work that can be done in parallel, the work may be wrapped up in a function
and queued with its arguments for later execution using a \texttt{spawn}.
Second, a programmer can use \texttt{spawn\_on} to spawn a task on a specific
core in the system or at the home core of a particular memory location. Third,
the programmer can invoke a parallel for loop, provided that the trip count is
known at loop entry. The programmer specifies a function pointer along with
start and end indices and an optional threshold to control parallel overhead.
\Grappa does {\em recursive decomposition} of iterations, similar to Cilk's
cilk\_for construct~\cite {cilkforimplementation}, and TBB's {\tt
parallel\_for}~\cite{intel_tbb}. It generates a logarithmically-deep tree of
tasks, stopping to execute the loop body when the number of iterations is
below the required threshold. Fourth, a programmer may want to execute an active message; that is, to run a
small piece of code on a particular core in the system without waiting for
execution resources to be available. Custom synchronization primitives, for example, execute this way as a function executed on the core where the data
lives. \Grappa provides the \texttt{call\_on} call for this purpose.
